---
title: "Germline classification"
output: github_document
---

```{r setup0, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(here)
library(yaml)
library(tidyverse)
library(ggh4x)
library(gt)
library(ggparty)
library(ggalluvial)
library(ggrepel)
library(patchwork)
config <- read_yaml("config.yaml")
extra_tumours <- config$extra_tumours
data_dir <- config$germline_classification_folder
source("scripts/plot_utils.R")
knitr::opts_chunk$set(root.dir = here(), comment = NA, echo = FALSE, message = FALSE, warnings = FALSE, include = TRUE, fig.align = 'center', fig.retina = 2, fig.path = "figures/", cache.path = "cache/")
```

With the hypothesis that if tumour HRD profile can inform us about the pathogenicity of a germline mutation, we need to first classifify individuals as BRCA1/2_positive or BRCA_negative groups based on their germline variants.

To stratify individuals, we first collect all germline mutations of each individual in 10 target genes including *BRCA1, BRCA2, PALB2, BARD1, BRIP1, RAD51C, RAD51D, CHEK2*, and *ATM*. The germline variant calls were obtain from our best-practice pipeline on HPC and each variant was annotated using `nanno` tool (better VEP from QIMR).

A set of unique SNVs and indels from all samples were collected for curation.

```bash
cat *.filtered.tsv | \
grep -v "#" | \
cut --complement -f 5| \
sort | uniq > BRCA.germline_snv.annotation.tsv
```

After curation, each variant was labelled as A, B, C, D, E, F class as below:


```{r}
variant_class <- LETTERS[1:6]
criteria <- c(
  "Pathogenic or hot VUS in HR genes other than BRCA1/2, and also ATM, CHEK2",
  "(cold) VUS in HR genes other than BRCA1/2, and also ATM, CHEK2",
  "Benign/Likely benign variants in HR genes other than BRCA1/2, and also ATM, CHEK2",
  "Pathogenic variants in BRCA1 or BRCA2",
  "VUS in BRCA1 or BRCA2",
  "Benign/Likely benign variants in BRCA1 and/or BRCA2"
)

num_of_variant <- read_tsv("../4.brca_classification/germline_mutation_classification.txt") |> count(Class_updated) |> pull(n)

tibble("Variant class" = variant_class, "Criteria" = criteria, "N"=num_of_variant) |> 
  gt() |> tab_footnote(footnote=md("*Genes included: BRCA1, BRCA2, PALB2, BARD1 ,BRIP1, RAD51C, RAD51D, CHEK2, ATM*")) |> as_raw_html()

```

Next, we went through each sample, and labelled them as BRCA1/2_positive or BRCA1/2_negative using [label_individual.py](scripts/label_individual.py). Furthermore, to avoid the noise from pathogenic variants of other genes, we excluded samples if P/LP variants of other genes (class A) or VUS of other genes(class B) were present in one. In addition, we separated individuals with VUS_BRCA (class E) variants but not others as the Additional group for further investigation. 

```bash
cut -f1,2,3,4,8,100 BRCA.germline_snvs.annotation_cf.final.txt > germline_mutation_classification.txt. # get chr, position, ref, alt, gene_name, (variant_class) columns in xlsx from CF.
for file in *.filtered.tsv; do 
  python label_individual.py $file >> sample_classification.txt
done
```

We merged with this classification with thier tumour HRD profiles and generate a big table ()[]: 

```{r}
germline_df <- read_tsv(paste0(data_dir, "sample_classification.txt"),
         col_names=c("sample_id","A","B","C","D","E","F","germline_classification","class_D_mutation","class_E_mutations","class_A_mutations","class_B_mutations")) 


chord_df    <- read_tsv("results/chord_hrdprediction_355.tsv") |> select(sample,p_hrd,p_BRCA1,p_BRCA2,hr_status,hrd_type)
hrdetect_df <- read_tsv("results/hrdetect_hrdprediction_355.tsv") |> select(sample, Probability)
sigs_df     <- read_tsv("results/signature_tools_lib_sbs.355.tsv") |> column_to_rownames("sample") |> (\(x) x/rowSums(x))() |> rownames_to_column("sample") |> select(sample,Signature_3, Signature_8)
hrdsum_df   <- read_tsv("results/hrdsum_355.tsv") |> dplyr::rename("HRDsum" = `HRD-sum`, "LOH"=HRD, "TAI"=`Telomeric AI`)

somatic_df <- inner_join(chord_df, hrdetect_df) |> 
  inner_join( sigs_df) |> 
  inner_join( hrdsum_df) |> 
  mutate(cohort=case_when(startsWith(sample,"MAGIC")~"MAGIC",
                          startsWith(sample,"FBC")~"Familial Breast",
                          startsWith(sample,"TCGA")~"TCGA-BRCA",
                          TRUE~"Q-IMPROvE"),
         tumour_id=ifelse(cohort %in% c("MAGIC","Q-IMPROvE"), str_split_i(sample,"\\.",i=2), sample),
         sample_id=ifelse(cohort %in% c("MAGIC","Q-IMPROvE"), str_split_i(sample,"\\.",i=1), sample)) |> 
  filter(!tumour_id  %in% extra_tumours) |> 
  select(-c(sample)) |> 
  select( cohort,sample_id,everything()) 
```


```{r}
qc_df <- read_tsv("results/purity_tmb.tsv")
df <- inner_join(qc_df, somatic_df, by="sample_id")  |> inner_join(germline_df) |> select(cohort, sample_id, tumour_id, everything())
#df |> write_tsv("results/brca_germline_clf_tumor_hrd.tsv")

head(df, n=5) |> gt() |> 
   tab_spanner(
    label = md("**CHORD results**"),
    columns = c(p_hrd:hrd_type)
  ) |> 
   tab_spanner(
    label = md("**HRDetect results**"),
    columns = c(Probability:Signature_8)
  ) |> 
  tab_spanner(
    label = md("**HRsum results**"),
    columns = c(LOH:HRDsum)
  ) |> 
   tab_spanner(
    label = md("**Germline classification**"),
    columns = c(A:class_B_mutations)
  ) |>
  sub_missing() |> 
  as_raw_html()
```


### A partition tree of the individual germline classification

```{r, warning=FALSE}
df <- read_tsv("results/brca_germline_clf_tumor_hrd.tsv") 
df <- df |> mutate(hr_status_combined=case_when(Probability>=0.7 & hr_status=="HR_deficient"~"HRD", 
                                      Probability<0.7 & hr_status=="HR_proficient" ~ "non-HRD", 
                                      TRUE~"Inconsistent" ) )


tree_df <- df |> select(A, B, C, D, C, E, F, sample_id, , hr_status_combined, germline_classification) |>  
  filter(!hr_status_combined=="Inconsistent") |> 
  mutate(has_A=ifelse(A>0, "yes", "no") |> factor(levels=c("no","yes")),
         has_B=ifelse(B>0, "yes", "no") |> factor(levels=c("no","yes")),
         has_C=ifelse(C>0, "yes", "no") |> factor(levels=c("no","yes")),
         has_D=ifelse(D>0, "yes", "no") |> factor(levels=c("no","yes")),
         has_E=ifelse(E>0, "yes", "no") |> factor(levels=c("no","yes")),
         has_F=ifelse(F>0, "yes", "no") |> factor(levels=c("no","yes"))) |> 
  select(has_A, has_B, has_C, has_D, has_C, has_E, has_F, germline_classification, sample_id, hr_status_combined) |> 
  as.data.frame()

sp_a <- partysplit(1L, breaks = 0)
sp_d <- partysplit(4L, breaks = 0)
sp_e <- partysplit(5L, breaks = 0)
sp_b <- partysplit(2L, breaks = 0)

sp_a <- partysplit(1L, index=c(1L,2L))
sp_d <- partysplit(4L, index=c(1L,2L))
sp_e <- partysplit(5L, index=c(1L,2L))
sp_b <- partysplit(2L, index=c(1L,2L))

pn <- partynode(1L, split = sp_a, kids = list(
  partynode(3L, split = sp_d, kids = list(
    partynode(5L, split = sp_e, kids = list(
      partynode(7L, split = sp_b, kids = list(
        partynode(9L, info = "BRCA1/2 negative"),
        partynode(8L, info = "Excluded VUS")
      )),
      partynode(6L, info = "BRCA1/2 additional")
    )),
    partynode(4L, info = "BRCA1/2 positive")
  )),
  partynode(2L, info = "Excluded P/LP")
))

tree <- party(pn,
            data = tree_df,
            fitted = data.frame(
              "(fitted)" = fitted_node(pn, data = tree_df),
              "(response)" = tree_df$hr_status_combined,
              check.names = FALSE),
            terms = terms(hr_status_combined ~ ., data = tree_df)
)

```

```{r, Figure2, fig.height=8, fig.width=10.6, warning=FALSE, fig.cap="This figure illustrates the method we classfied each individual as different germline group"}
p_tmp <- tree_df |> 
  ggplot(aes(x=hr_status_combined, fill=hr_status_combined)) + 
  geom_bar() + 
  scale_fill_manual(values=c("#0073C2","#EFC000")) +
  labs(fill="") + theme(legend.position = "bottom")
p_legend <- ggpubr::get_legend(p_tmp)

p1 <- ggparty(tree) +
  geom_edge() +
  geom_edge_label() +
  geom_node_splitvar(fontface = "bold") +
  geom_node_plot(gglist = list(geom_bar(aes(x = hr_status_combined, fill = hr_status_combined)),
                               geom_text(stat='count', aes(x=hr_status_combined,label=..count..), vjust=0, size=2.6),
                               theme_bw(),
                               theme(axis.text.x = element_blank(), 
                                     axis.ticks.x = element_blank(), 
                                     panel.grid = element_blank()),
                               scale_fill_manual(values=c("#0073C2","#EFC000")),
                               ylab("Count"),
                               xlab(""))) +
  geom_node_label(aes(label = paste0(info, ", N = ", nodesize)),
                  fontface = "bold",
                  ids = "terminal",
                  size = 3,
                  # 0.01 nudge_y is enough to be above the node plot since a terminal
                  # nodeplot's top (not center) is at the node's coordinates.
                  nudge_y = 0.01) 

p1/p_legend + plot_spacer() + plot_layout(heights = c(11,-1.5,1))
```

Furthermore, depending on whether the mutation is present in BRCA1 or BRCA2, we categorized additional mutations as either additional BRCA1 or additional BRCA2, and BRCA1/2-positive cases as either positive BRCA1 or positive BRCA2.

**The number of samples in each detailed category**

```{r, snv_germline_classification_combined,  fig.width=9, fig.height=4.8}
group_labeller <- c("additional_brca1"="BRCA1 VUS","additional_brca2"="BRCA2 VUS", "exclude"="Exclude","negative_brca1/2"="BRCA1/2 negative", "positive_brca1"="BRCA1 positive", "positive_brca2"="BRCA2 positive")

df |> filter(hr_status_combined!="Inconsistent") |> 
  ggplot(aes(x=hr_status_combined, fill=hr_status_combined)) +
  geom_bar() + 
  geom_text(stat='count', aes(x=hr_status_combined,label=..count..), vjust=-.2, size=4) +
  facet_grid(~germline_classification, labeller = labeller(germline_classification=group_labeller)) +
  scale_fill_manual(values=c("#0073C2","#EFC000")) +
  labs(x="",y="Count", fill="HR status") + ylim(0,150) +
  theme_Publication() +
  theme(strip.text = element_text(size=10))
```



We also examed structrual variations (SVs) in samples using `DELLY`. The SVs in the 10 genes we listed before were extracted for curation, and those potential pathogenic or hot VUS in other genes excpete BRCA1/2 (class A), pathogenic variants in BRCA1/2 (class D) and VUS in BRCA1/2 (class E) were visually checked in IGV. 

Eventually, this results in  7 SVs:

```{r}
read_tsv("../4.brca_classification/sv/real_sv.txt", col_names = c("chrom1","pos1","chrom2","pos2", "ID", "SV_type", "SV_len", "gene1", "gene2", "frequency","cohort","class")) |> 
  gt() |> as_raw_html()
```

We used [extract_ind_with_pathSV.py](scripts/extract_ind_with_pathSV.py) to get individuals carries these SVs in each cohort.

```bash
python extract_ind_with_pathSV.py $cohort | datamash -s -g1 collapse 2 > ${cohort}.sv.tsv
```

After integrating SV results in individuals classification, 6 samples were affected by these real SVs, we then updated the germline classfication for these individuals 

```{r}
sv_df <- read_tsv("../4.brca_classification/sv/sample_pathSV.real.tsv", col_names = c("sample_id","pathSV"))
sv_df |> gt() |> as_raw_html()
df2 <- df |> left_join(sv_df) |> 
  mutate(SV_A = map_dbl(str_split(pathSV, ","), ~ sum(str_detect(., "^A"))),
         SV_D = map_dbl(str_split(pathSV, ","), ~ sum(str_detect(., "^D"))),
         SV_E = map_dbl(str_split(pathSV, ","), ~ sum(str_detect(., "^E")))) |> 
  replace_na(list(SV_A=0, SV_D=0, SV_E=0)) |> 
  mutate(A=ifelse(SV_A>0, A+SV_A, A),
         D=ifelse(SV_D>0, D+SV_D, D),
         E=ifelse(SV_E>0, E+SV_E, E)) |> 
  select(-c(SV_A:SV_E))

affected_samples <- sv_df |> pull(sample_id)
tmp_df <- df2 |> filter(sample_id %in% affected_samples) |> select(cohort,sample_id, hr_status_combined, A, B, C, D, E, F, germline_classification, pathSV) |> 
  mutate(updated_germline_classification=case_when(
    A>0 ~ "exclude",
    D>0 & grepl("BRCA1", pathSV) ~ "positive_brca1",
    D>0 & grepl("BRCA2", pathSV) ~ "positive_brca2",
    TRUE~ germline_classification
  )) 

tmp_df |> select(-pathSV) |>  gt() |> as_raw_html()


result_df <- df2 |> left_join(tmp_df |> select(sample_id, updated_germline_classification)) |> 
  mutate(updated_germline_classification=ifelse(is.na(updated_germline_classification), germline_classification, updated_germline_classification)) 

result_df |> write_tsv("results/brca_germline_clf_tumor_hrd.add_sv.tsv")
```

### Partitioin tree after integrating SVs.

```{r Figure2.1, fig.height=8, fig.width=10.6, warning=FALSE}
tree_df2 <- df2 |> select(A, B, C, D, C, E, F, sample_id, , hr_status_combined) |>  
  filter(!hr_status_combined=="Inconsistent") |> 
  mutate(has_A=ifelse(A>0, "yes", "no") |> factor(levels=c("no","yes")),
         has_B=ifelse(B>0, "yes", "no") |> factor(levels=c("no","yes")),
         has_C=ifelse(C>0, "yes", "no") |> factor(levels=c("no","yes")),
         has_D=ifelse(D>0, "yes", "no") |> factor(levels=c("no","yes")),
         has_E=ifelse(E>0, "yes", "no") |> factor(levels=c("no","yes")),
         has_F=ifelse(F>0, "yes", "no") |> factor(levels=c("no","yes"))) |> 
  select(has_A, has_B, has_C, has_D, has_C, has_E, has_F, sample_id, hr_status_combined) |> 
  as.data.frame()


tree2 <- party(pn,
            data = tree_df2,
            fitted = data.frame(
              "(fitted)" = fitted_node(pn, data = tree_df2),
              "(response)" = tree_df2$hr_status_combined,
              check.names = FALSE),
            terms = terms(hr_status_combined ~ ., data = tree_df2)
)


p2 <- ggparty(tree2) +
  geom_edge() +
  geom_edge_label() +
  geom_node_splitvar(fontface = "bold") +
  geom_node_plot(gglist = list(geom_bar(aes(x = hr_status_combined, fill = hr_status_combined)),
                               geom_text(stat='count', aes(x=hr_status_combined,label=..count..), vjust=0, size=2.6),
                               theme_bw(),
                               theme(axis.text.x = element_blank(), 
                                     axis.ticks.x = element_blank(), 
                                     panel.grid = element_blank()),
                               scale_fill_manual(values=c("#0073C2","#EFC000")),
                               ylab("Count"),
                               xlab(""))) +
  geom_node_label(aes(label = paste0(info, ", N = ", nodesize)),
                  fontface = "bold",
                  ids = "terminal",
                  size = 3,
                  # 0.01 nudge_y is enough to be above the node plot since a terminal
                  # nodeplot's top (not center) is at the node's coordinates.
                  nudge_y = 0.01)

p2/p_legend + plot_spacer() + plot_layout(heights = c(11,-1.5,1))
```

Lastly, we add somatic mutation information as well using [add_somatic_mutations.py](scripts/add_somatic_mutations.py). The results include any somatic mutations in the 10 genes and the copy number of each gene.

```bash
python add_somatic_mutations.py brca_germline_clf_tumor_hrd.add_sv.tsv > brca_germline_clf_tumor_hrd.add_sv.add_somatic.tsv
```

```{r}
read_tsv("results/brca_germline_clf_tumor_hrd.add_sv.add_somatic.tsv") |> head(5) |> 
  gt() |> 
  sub_missing() |> 
  as_raw_html()
```


Previous tree plot didn't include individuals have inconsistent HRD prediction from HRDetect and CHORD. 
If just ignore the tumour HRD type, all 350 individuals were classified as different groups by either carries germline variant in category A, E, B or D.


```{r,fig.height=8, fig.width=15.6, warning=FALSE }
tree_df3 <- df2 |> select(A, B, D, C, E, sample_id) |>  
  mutate(Carries_A=ifelse(A>0, "yes", "no") |> factor(levels=c("no","yes")),
         Carries_B=ifelse(B>0, "yes", "no") |> factor(levels=c("no","yes")),
         Carries_C=ifelse(C>0, "yes", "no") |> factor(levels=c("no","yes")),
         Carries_D=ifelse(D>0, "yes", "no") |> factor(levels=c("no","yes")),
         Carries_E=ifelse(E>0, "yes", "no") |> factor(levels=c("no","yes")),
         Carries_F=ifelse(F>0, "yes", "no") |> factor(levels=c("no","yes")),) |> 
  select(Carries_A, Carries_B, Carries_C, Carries_D, Carries_E,Carries_F, sample_id) |> 
  as.data.frame()


tree3 <- party(pn,data = tree_df3)
ggparty(tree3) +
  geom_edge() +
  geom_edge_label() +
  geom_node_splitvar(fontface = "bold") +
  geom_node_label(aes(label = paste0(info, ", N = ", nodesize)),
                  fontface = "bold",
                  ids = "terminal",
                  size = 2.5) +
  theme_void()
```



## The final Figure1.

Figure one is about the germline classification

```{r germline_classification_alluvium_plot, fig.width=8.6, fig.height=9, warning=FALSE }
allu_df <-  result_df |> select(cohort, A, B, C, D, C, E, F, sample_id, updated_germline_classification) |>  
  mutate(has_A=ifelse(A>0, "yes", "no") |> factor(levels=c("yes","no")),
         has_B=ifelse(B>0, "yes", "no") |> factor(levels=c("yes","no")),
         has_C=ifelse(C>0, "yes", "no") |> factor(levels=c("yes","no")),
         has_D=ifelse(D>0, "yes", "no") |> factor(levels=c("yes","no")),
         has_E=ifelse(E>0, "yes", "no") |> factor(levels=c("yes","no")),
         has_F=ifelse(F>0, "yes", "no") |> factor(levels=c("yes","no"))) |> 
  select(cohort, has_A, has_B, has_C, has_D, has_C, has_E, has_F, sample_id, updated_germline_classification) |> 
  mutate(germline_classification=ifelse(startsWith(updated_germline_classification,"additional"),"additional",updated_germline_classification)) |> 
  select(-updated_germline_classification) |> 
  group_by(cohort, germline_classification) |> mutate(n=n()) 

need_repel <- c("Q-IMPROvE", "positive_brca1","positive_brca2", "additional")
alluvial_labeller <- c("BRCA1/2 VUS ","Exclude","BRCA1/2 negative", "BRCA1 positive", "BRCA2 positive", stringr::str_wrap("Familial Breast", 5),"MAGIC","TCGA-BRCA","Q-IMPROvE", "Yes","No")
names(alluvial_labeller) <- c("additional","exclude","negative_brca1/2","positive_brca1","positive_brca2","Familial Breast","MAGIC","TCGA-BRCA","Q-IMPROvE", "yes","no")



allu_df$cohort <- factor(allu_df$cohort, levels = cohort_levels)
allu_df$germline_classification <- factor(allu_df$germline_classification, levels=c("positive_brca1","positive_brca2","exclude","additional","negative_brca1/2"))

group_color_mapping <- c("additional"="#0099B4", "exclude" = "#FDAF91", "negative_brca1/2"="#ADB6B6", "positive_brca1" = "#00468B", "positive_brca2"="#ED0000")
group_colour_mapping <- group_color_mapping # ZJ: super weired bug if not having this : object 'group_colour_mapping' not found, I'm not using it anywhere.

p_allu <- ggplot(allu_df |> arrange(cohort),aes(axis6 = cohort, axis5 = has_A, axis4 = has_D, axis3= has_E, axis2=has_B, axis1=germline_classification,
                                                y = n)) +
  scale_x_discrete(limits = c("cohort", "has_A", "has_D", "has_E", "has_B","germline_classification") |> rev() , 
                   labels=c(stringr::str_wrap("Germline classification", 5),"Carries B?","Carries E?","Carries D?","Carries A?","Cohort"), expand = c(.15,.05)) +
  scale_y_continuous(expand = c(0.01, 0.0001)) +  
  geom_alluvium(aes(fill=germline_classification),reverse = FALSE) + scale_fill_manual(values = group_color_mapping) +
  geom_stratum(width = 0.4, reverse = FALSE, aes(fill=ifelse(after_stat(stratum) %in% names(group_color_mapping), as.character(after_stat(stratum)), NA))) + 
  scale_fill_manual(values=group_color_mapping, na.value = "white") +
  geom_text(stat = "stratum", aes(label=ifelse(after_stat(stratum) %in%c("yes","no"), alluvial_labeller[as.character(after_stat(stratum))], "")), size=2.8, reverse = FALSE)+
  geom_text(stat = "stratum", aes(label=ifelse(after_stat(stratum) %notin% c(need_repel, "yes","no"), alluvial_labeller[as.character(after_stat(stratum))], "")), size=3.6, reverse = FALSE)+
  geom_text_repel(stat = "stratum", aes(label = ifelse(after_stat(stratum) %in% need_repel & after_stat(x) == 6, as.character(after_stat(stratum)), "")), size = 3.6,
                  direction = "y",
                  nudge_x = .35,
                  reverse = FALSE, segment.color="darkgrey") +
  geom_text_repel(stat = "stratum", aes(label = ifelse(after_stat(stratum) %in% need_repel & after_stat(x) == 1, alluvial_labeller[as.character(after_stat(stratum))], "")), size = 3.6,
                  direction = "y",
                  nudge_x = -.45,
                  reverse = FALSE, segment.color="darkgrey") +
  theme_void() + coord_flip() + 
  theme(legend.position = "none",
        axis.text.y= element_text())


p_germline_classification_by_cohort <- result_df |> 
  mutate(germline_classification=ifelse(startsWith(updated_germline_classification,"additional"),"additional",updated_germline_classification)) |> 
  count(cohort,germline_classification) |> 
  complete(cohort,germline_classification) |> 
  mutate(n = replace_na(n, 0), cohort=factor(cohort, levels=cohort_levels)) |> 
  mutate(germline_classification = factor(germline_classification, levels=c("positive_brca1","positive_brca2","exclude","additional","negative_brca1/2"))) |> 
  ggplot(aes(x = cohort, y = n, fill = germline_classification)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  geom_text(aes(label = n),position = position_dodge(width = 0.7),vjust = -0.5, size=2.5) +
  theme_Publication() +
  scale_x_discrete(limits=cohort_levels, labels=c("Familial Breast","TCGA-BRCA","MAGIC","Q-IMPROvE")) +
  scale_fill_manual( labels=alluvial_labeller,values = group_color_mapping) + 
  labs(x = "", y = "Count", fill = "") +
  scale_y_continuous(expand = expansion(mult = c(.2, 0.16))) +
  theme(axis.title.y=element_text(vjust=-12), legend.text = element_text(size=10), legend.title = element_text(size=10))


p_allu / p_germline_classification_by_cohort + plot_layout(heights = c(0.88,0.12))
```

**The classification process of 350 inviduals and the number of individuals labelled as different groups in 4 cohorts**
